Animacy Detection in Stories

### 2014-12-10 paper/paper.tex
updates

For almost all species in the world, the capacity to distinguish
animate objects from inanimate objects is essential to their
survival. Those objects could be prey, for example, or predators or
mates. The fundamental nature of the distinction between animate and
inanimate appears in the early acquisition of the division
\cite{opfer:02}. Children of less than 6 months old are well able to
make the division. Recent brain research shows that the distinction
appears in the organization of the brain (e.g.~\cite{gao:12}). For some
researchers this provides evidence for the idea that the division
between animate and inanimate is an innate part of how we see the
world.

A common view supported by philosophers such as Daniel Dennett is that
animacy and aliveness are better to be treated as epistemological
stances rather than fixed states in the world \cite{dennett:96}. Not
ineffable qualia but behavioral capacity defines our stance towards
objects. Depending on whether people \textit{think} that an object is
animate, they utilize different cognitive strategies to explain and
predict the actions of those objects. The following opening of a
well-known story will make this point more clear:

@@

Given the fundamental nature of the distinction between animate and
inanimate, it is perhaps not too surprising that it has proven to be
useful in a variety of natural language processing tasks dealing with
e.g.\ anaphora resolution and dependency parsing
\cite{orasan:07,lee:13,ovr:niv:07}. Existing methods to automatically
label natural language for animacy are usually rule based or apply
machine learning techniques or make a combination of both. Common to
most approaches is that they make use of semantic lexicons with
information about animacy and syntactic cues in a text. Both feature
types are relatively costly to obtain as they require large
vocabularies or syntactic parsing systems, which are besides for a
handful of western languages, often not available.

@@

The rest of this paper is structured as follows. We will start with an
short overview of existing techniques to automatically label texts for
animacy (\S\ref{sec:previous-work}). After a description of the corpus
used in our study and how the annotations of the corpus have been
established (\S\ref{sec:data}), we will give an account of our
computational models in Section \ref{sec:models}. We report on the
empirical results in Section \ref{sec:results}. The last section
offers our conclusions and possible directions for future research.

@@

In most of the literature on automatic animacy detection, people make
a binary distinction between animate and inanimate. \cite{bowman:12}
for example, define objects to be animate if they are alive and have
the ability to move under their own will. \cite{orasan:07} define
animacy in the context of anaphora resolution: ``if its referent can
also be referred to using one of the pronouns he, she, him, her, his,
hers, himself, herself, or a combination of such pronouns
(e.g. his/her)''. Not only are some parts of these definitions somewhat
circular, we believe that they are both essentialistic and not
particularly suited for the rich and wondrous entities that live in
the realm of stories. We choose to define animacy in terms of
Dennett's intentional stance which comes down to the question whether
``you decide to treat the object whose behavior is to be predicted as
a rational agent'' \cite[pp. 17]{dennett:96}.

@@

\cite{ovrelid:04,ovrelid:05,ovrelid:06,ovrelid:08,ovrelid:09} present
a number of animacy classifiers that make use of syntactic and
morphological features. Some of the features used are: how often has a
noun been analyzed as `subject' or `object'? How often does a noun
occur in a passive by-phrase? How often does a noun occur as a
subject, immediately followed by animate personal pronouns or
inanimate personal pronouns?  These features are then aggregated for
each lemma after which a machine learning system (decision tree or
$k$-nearest neighbor classifier) is trained. In \cite{bowman:12} a
similar approach is presented. In this study a maximum entropy
classifier is trained on the basis of three feature types: (1)
bag-of-words with and without their corresponding Part-of-Speech tags,
(2) internal syntactic features such as the syntactic head and (3)
external syntactic features that describe the dependency relation of a
noun to a verb (i.e.\ subject relation, object relation etc.)  This is
the only study that makes use of a fully labeled corpus for animacy.
Not so much concerned with animacy detection, \cite{karsdorp:12}
attempt to extract the cast (i.e.\ all characters) from a
story. Similar to \cite{bowman:12} they rely on dependency tags to
extract the subjects of direct and indirect speech.

@@

% commentary on above approaches
There are a number of problems with the approaches mentioned
above. First, almost all of them heavily rely on costly,
linguistically informed features derived from lexical-semantic
databases or syntactic parsing. For most languages in the world,
however, we cannot rely on these resources. Second, animacy detection
is often seen as a useful feature for a range of natural language
processing techniques, such as anaphora resolution and syntactic
parsing. The mutual dependence of these techniques and animacy
detection, however, leads to a chicken and egg situation. Another
problem with the approaches above is that they are lemma based which
means that the models are generally insensitive to different usages of
a particular word. A pancake is typically not an animate entity, but,
as we have seen, whether it is animate or not is only to be determined
in relation to the story in which it appears. We therefore need to
develop a dynamic, data driven system that is token based and does not
rely too much on static and fixed world knowledge.

@@

Our corpus consists of 74 Dutch stories from the collection
\textit{Volkssprookjes uit Nederland en Vlaanderen} compiled by
\cite{sinninghe:78}. The collection contains retellings of popular
stories from the Netherlands and Flanders, such as \textit{The Bremen
  Town Musicians} (ATU 130) and \textit{The Table, the Ass, and the
  Stick } (ATU 563). Some of the less well-know stories are
\textit{Cock, Hen, Duck, Pin, and Needle on a Journey} (AT 0210) and
(ATU 780) \textit{The Singing Bone}. A digital version of the
collection is available in the Dutch Folktale Database from the
Meertens Institute (corpus SINVSUNV.20E).\footnote{See
  \url{http://www.verhalenbank.nl}} We chose this collection because
of its homogeneity while it still displays enough diversity: (1) all
tales are written in standard Dutch, (2) the stories are edited by the
same editor, (3) the tales have a comparable length and (4) the
collection contains tales from a number of different genres, such as
fairy tales, legends, riddles etc.

We assigned to each word in the collection a binary label `animate' or
`inanimate'. 11542 words were labeled as animate and 62926 words as
inanimate. The vocabulary size is 5549 with 743 unique animate words
and 5011 unique inanimate words. The annotations consist of 4627
animate words that have been tagged as noun or name against 6878
inanimate words.

@@

@@

We formulate the problem of animacy detection as a classification
problem where the goal is to assign to each word in a document a label
indicating whether it represents an animate or an inanimate word.

@@

Animate words are far outnumbered by inanimate words in our
collection. Reporting accuracy scores would therefore provide skewed
results too much in favor of the majority category. The rarity of
animate words makes evaluation measures such as the well-known $F1$-score
more appropriate. For all experiments we report on the precision,
recall and $F1$-score \cite{rijsbergen:79}. In most of the literature on
animacy detection results are only given for the classification of
nouns or noun phrases. We will report the results for all words in a
text and on those specifically for nouns separately.

In real world applications, an animacy detection system will most
likely be faced with completely new texts instead of single words. It
is therefore important to construct a training and test procedure in
such a way that it mimics this situation as closely as possible. If we
would make a random split of say 80\% of the data for training and
20\% for testing on the word level, we run the risk of mixing training
data with test data making it too easy for a system to rely on words
it has seen from the same text. \cite{bowman:12} fall into this trap
by making a random split in their data on the sentence level. It is
very likely that in their setup sentences from the same document are
present in both the training data and the test data which makes their
evaluation unrealistic. To circumvent this problem, we train our model
on 80\% of the stories (n=59) and test on the remaining 15 stories.

@@

\subsubsection{Morphological Features}
For each word we extract its Part-of-Speech tag. For reasons of
comparability we choose to use the tags as provided by the output of
the syntactic parser Alpino, instead of a more specialized
Part-of-Speech tagger. Again, we take a sliding window approach and
extract the Part-of-Speech tags for three word left of the focus word,
three word right of the focus word and the tag of the focus word
itself.

@@

@@

We have trained word embeddings with 300 dimensions using the popular
skip-gram architecture \cite{mikolov:13} on the Dutch corpus of COW
(COrpora from the Web).\footnote{The word embedding vectors have been
  published online at XXX.} COW is a collection of linguistically
processed web corpora for English, Dutch, Spanish, French, Swedish and
German \cite{schaefer:12}. The 2014 Dutch corpus contains 6.8 billion
word tokens. The idea behind using the word embeddings is that
similarities between animate words can be estimated by inspecting the
context in which they occur. The hypothesis is that the word
embeddings of an animate word are more similar to those of other
animate words than than to those of inanimate words. To give an
example, the 5 nearest neighbors of \textit{Assepoester}
(`Cinderella') are: \textit{Sneeuwwitje} (`Snow White') (sim = 0.807),
\textit{Roodkapje} (`Little Red Riding Hood') (sim = 0.747),
\textit{Doornroosje} (`Sleeping Beauty') (sim = 0.715),
\textit{Repelsteeltje} (`Rumpelstiltskin') (sim = 0.669) and
\textit{Alladin} (0.645). In contrast, the 5 most similar words to
\textit{boek} (`book') are: \textit{boekje} (`book-\textsc{dim}')
(sim=0.8), \textit{kinderboek} (`children's book') (0.757),
\textit{stripboek} (`comic') (0.713), \textit{kortverhaal} (`short
story') (0.709), \textit{jeugdboek} (`youth book') (0.706).

@@

@@

In table \ref{tab:results-all} we present the results for all nine
models on the complete data set. For each model we report the
precision, recall and $F1$-score for the animate words and the
inanimate words. All models perform well on classifying inanimate
words. However, since this is the majority class, it is far more
interesting to compare the performance of the models on the animate
instances. It is interesting to observe that the `simple' $n$-gram
word model already performs rather well. Adding more features, like
Part-of-Speech, lemmata, etcetera, only has a positive impact on the
recall of the model leaving the precision untouched. As can be
observed from the table, the rather expensive dependency features show
barely any improvement. The model that only uses word embedding
features, is one of the best performing models. This was slightly
unexpected, since this is the model in which features are context
insensitive. However, the model that combines the word embeddings with
the word $n$-gram features outperforms the other linguistically
informed models that make use of lemmata, Part-of-Speech tags or
dependency information. The best performance is achieved by the model
that combines the word features, Part-of-Speech tags and the word
embeddings with an $F1$-score of 0.93 for animate words and 0.99 on
inanimate words. Adding more features does not result in any more
performance gain.

@@

The approach taken in this paper to create a model for animacy
classification using linguistically uninformed features proves to be
successful. We compared the performance of linguistically informed
models (using features such as Part-of-Speech and dependency tags) to
models that make use of lower dimensional representations of the
data. With the exception of the model that solely makes use of these
representations, all models benefit from adding these features. The
model that requires the least linguistic information (word $n$-grams
plus word embeddings) outperforms all linguistically informed
models. The best results are reported by the model that combines word
$n$-grams with Part-of-Speech $n$-grams and word embeddings.

>>>>

For almost all species in the world, the capacity to distinguish
animate objects from inanimate objects is essential to their
survival. Those objects could be prey, for example, or predators or
mates. The fundamental nature of the distinction between animate and
inanimate is reflected in the fact that the division is acquired very early in life.
\cite{opfer:02}. Children of less than 6 months old are well able to
make the division. Moreover, recent brain research shows that the distinction
appears in the organization of the brain (e.g.~\cite{gao:12}). For some
researchers this provides evidence for the idea that the division
between animate and inanimate is an innate part of how we see the
world.
%hoe sluit dit aan op de eerdere alinea? lijkt alsof je jezelf tegenspreekt namelijk: als het een standpunt is, waarom zien we dat dan niet in de hersens?%
A common view supported by philosophers such as Daniel Dennett is that
animacy and aliveness are better to be treated as epistemological
stances rather than fixed states in the world \cite{dennett:96}: not
ineffable qualia but behavioral capacity defines our stance towards
objects. Depending on whether people \textit{think} that an object is
animate, they utilize different cognitive strategies to explain and
predict the actions of those objects. Th following opening of a
well-known story will make this point more clear:

@@

Given the fundamental nature of the distinction between animate and
inanimate, it is perhaps not too surprising that it has proven to be
useful in a variety of natural language processing tasks dealing with
e.g.\ anaphora resolution and dependency parsing
\cite{orasan:07,lee:13,ovr:niv:07}. Existing methods to automatically
label natural language for animacy are usually either rule based or apply
machine learning techniques, or make a combination of both. Common to
most approaches is that they make use of semantic lexicons with
information about animacy and syntactic cues in a text. Both feature
types are relatively costly to obtain as they require large
vocabularies or syntactic parsing systems, which, with the exception of a few languages, are not readily available.

@@

The rest of this paper is structured as follows. We will start with a
short overview of existing techniques for automatically labeling animacy in texts (\S\ref{sec:previous-work}). After a description of the corpus
used in our study and how the annotations of the corpus have been
established (\S\ref{sec:data}), we will give an account of our
computational models in Section \ref{sec:models}. We report on the
empirical results in Section \ref{sec:results}. The last section
offers our conclusions and possible directions for future research.

@@

%ik wil eigenlijk meer uitleg in deze paragraaf: 
In most of the literature on automatic animacy detection, a static, binary distinction is made
between animate and inanimate. \cite{bowman:12}
for example, define objects as animate if they are alive and have
the ability to move under their own will. \cite{orasan:07} define
animacy in the context of anaphora resolution: something is animate ``if its referent can
also be referred to using one of the pronouns he, she, him, her, his,
hers, himself, herself, or a combination of such pronouns
(e.g. his/her)''. However, some parts of these definitions are somewhat
circular and essentialistic. More importantly, they are not
particularly applicable to the rich and wondrous entities that live in
the realm of stories. As our example above showed, it is possible for an otherwise as inanimate labeled object to attain animacy. We therefore choose to define animacy in terms of
Dennett's intentional stance, which is more dynamic, and which ultimately comes down to the question whether
``you decide to treat the object whose behavior is to be predicted as
a rational agent'' \cite[pp. 17]{dennett:96}.

@@

\cite{ovrelid:04,ovrelid:05,ovrelid:06,ovrelid:08,ovrelid:09} present
a number of animacy classifiers that make use of syntactic and
morphological features. These features include the frequency of analysis of the noun as `subject' or `object', the frequency of the occurence of a noun in a passive $by$-phrase, and the frequency of the noun as a subject followed by either animate personal pronouns or
inanimate personal pronouns. These features are then aggregated for
each lemma after which a machine learning system (decision tree or
$k$-nearest neighbor classifier) is trained. In \cite{bowman:12} a
similar approach is presented. In this study a Maximum Entropy
classifier is trained on the basis of three feature types: (1)
bag-of-words with and without their corresponding Part-of-Speech tags,
(2) internal syntactic features such as the syntactic head and (3)
external syntactic features that describe the dependency relation of a
noun to a verb (i.e.\ subject relation, object relation etc.)  This is
the only study that makes use of a fully labeled corpus for animacy.
Not so much concerned with animacy detection, \cite{karsdorp:12}
attempt to extract the cast (i.e.\ all characters) from a
story. Similar to \cite{bowman:12} they rely on dependency tags to
extract the subjects of direct and indirect speech.

@@

% commentary on above approaches
The approaches mentioned above present us with a number of problems. First, almost all of them heavily rely on costly,
linguistically informed features derived from lexical-semantic
databases or syntactic parsing. For most languages in the world,
however, we cannot rely on these resources, either because they do not exist, or because %syntactic parsers have too low results%. 
Second, animacy detection is often seen as a useful feature for a range of natural language
processing techniques, such as anaphora resolution and syntactic
parsing. The mutual dependence of these techniques and animacy
detection, however, leads to a chicken and egg situation. Another
problem with the approaches above is that they are lemma-based, which
means that the models are generally insensitive to different usages of
a particular word. A pancake is typically not an animate entity, but,
as we have seen, whether it is animate or not may depend on the story in which it appears, and even within the story the animacy may change. We therefore need to
develop a dynamic, data driven system that is token based and does not
rely too much on static and fixed world knowledge.

@@

To develop this dynamic, data driven system we used  a corpus of Dutch folktales. The corpus consists of 74 Dutch stories from the collection
\textit{Volkssprookjes uit Nederland en Vlaanderen} compiled by
\cite{sinninghe:78}. The collection is made up of Dutch retellings of popular stories, including such tales as \textit{The Bremen
  Town Musicians} (ATU 130) and \textit{The Table, the Ass, and the
  Stick } (ATU 563), as well as lesser-known stories such as
\textit{The Singing Bone} (ATU 780) and \textit{Cock, Hen, Duck, Pin, and Needle on a Journey} (AT 0210). This last story is again a clear example of otherwise inanimate objects being animate in a folk story, as is concerns the adventures of several household items, such as a pin, a hackle, an egg, and a whetstone. A digital version of the
collection is available in the Dutch Folktale Database from the
Meertens Institute (corpus SINVSUNV.20E).\footnote{See
\url{http://www.verhalenbank.nl}} Using a single collection for our corpus presents us with a helpful homogeneity with regard to the editor, length and language use. On the other hand, the collection displays diversity as well in the choice of the stories, which entail fairy tales, legends, and sagas.
%getallen nakijken%
%moet je de labels hier al wel noemen? want dat lijkt me eerder een resultaat dan een pre-processing stap toch?%

The corpus consists of 74504 words, from 5549 unique words. All words or tokens were assigned a binary label `animate' or `inanimate', where all words that were not animate (e.g. determiners, adjectives, etc.) were labelled as inanimate. This resulted in 11542 animate tokens of 743 types, and 62926 inanimate tokens from 5011 unique inanimate words. A selection of words, such as the abovementioned pancake and egg, was labeled as both animate and inanimate, depending on the context. The distribution of the nouns is striking: of the animate tokens, 4627 (40,1\%) are nound, while only 6878 of the inanimate tokens (11 \%) are nouns. This shows that being a noun is already somewhat of an indication for animacy.

% voorbeeld met gloss 4 regels (NL, ani/inani/Engels/vrij vertaling)

@@

% gaat nu de volgorde niet fout? Eerst tokenizeren en parsen, en dan animate/inanimate toekennen?%

@@

We formulate the problem of animacy detection as a classification
problem where the goal is to assign a label at word level, rather than at lemma level,
indicating whether it represents an animate or an inanimate word.

@@

Animate words are far outnumbered by inanimate words in our
collection (see (\S\ref{sec:data})). Reporting accuracy scores would therefore provide skewed
results, strongly favoring the majority category. The relative rarity of
animate words makes evaluation measures such as the well-known $F1$-score
more appropriate. For all experiments we report on the precision,
recall and $F1$-score \cite{rijsbergen:79}. Also, while in most of the literature on animacy detection results are only presented for the classification of nouns or noun phrases, we will present these and also report on the results for all words in a
text.

In real world applications, an animacy detection system will most
likely be faced with completely new texts instead of single words. It
is therefore important to construct a training and test procedure in
such a way that it mimics this situation as closely as possible. If we
would, for example, make a random split of 80\% of the data for training and
20\% for testing on the word level, we run the risk of mixing training
data with test data, thereby making it too easy for a system to rely on words
it has seen from the same text. \cite{bowman:12} fall into this trap
by making a random split in their data on the sentence level. It is
very likely that in their setup sentences from the same document are
present in both the training data and the test data, making their
evaluation unrealistic. To circumvent this problem, we split at the story level, using 80\% of the stories (n=59) as training material and the remaining 15 stories as test.
%dat is de innovatie right? dat het niet gaat om woord of zin maar om de hele tekst%

@@

\subsubsection{Morphological Features}
For each word we extract its part-of-speech tag. For reasons of
comparability we choose to use the tags as provided by the output of
the syntactic parser Alpino, instead of a more specialized
part-of-speech tagger. Again, we take a sliding window approach and
extract the part-of-speech tags for three words left and right of the focus word, as well as the tag of the focus word itself.

@@

%is n hier weer 3? Je specificeert dat niet%

@@

%je moet een of meerdere bronnen noemen hier, by "recently there has been interest"%

We have trained word embeddings with 300 dimensions using the popular
skip-gram architecture \cite{mikolov:13} on the Dutch corpus of COW
(COrpora from the Web).\footnote{The word embedding vectors have been
  published online at XXX.} COW is a collection of linguistically
processed web corpora for English, Dutch, Spanish, French, Swedish and
German \cite{schaefer:12}. The 2014 Dutch corpus contains 6.8 billion
word tokens. The idea behind using the word embeddings is that
similarities between animate words can be estimated by inspecting the
context in which they occur. From this follows for example that the word
embeddings of an animate word are more similar to those of other
animate words, as opposed to the embeddings of inanimate words. To give an
example, the 5 nearest neighbors of \textit{Assepoester}
(`Cinderella') are: \textit{Sneeuwwitje} (`Snow White') (sim = 0.807),
\textit{Roodkapje} (`Little Red Riding Hood') (sim = 0.747),
\textit{Doornroosje} (`Sleeping Beauty') (sim = 0.715),
\textit{Repelsteeltje} (`Rumpelstiltskin') (sim = 0.669) and
\textit{Alladin} (0.645). In contrast, the 5 most similar words to
\textit{boek} (`book') are: \textit{boekje} (`book-\textsc{dim}')
(sim=0.8), \textit{kinderboek} (`children's book') (0.757),
\textit{stripboek} (`comic') (0.713), \textit{kortverhaal} (`short
story') (0.709), \textit{jeugdboek} (`youth book') (0.706).

@@

In table \ref{tab:results-all} we present the results for all nine
models on the complete data set. For each model we report the
precision, recall and $F1$-score for the animate words and the
inanimate words.

@@

All models perform well on classifying inanimate
words. However, since this is the majority class, it is far more
interesting to compare the performance of the models on the animate
instances. It is interesting to observe that the `simple' $n$-gram
word model already performs rather well. Adding more features, like
Part-of-Speech, lemmata, etcetera, only has a positive impact on the
recall of the model, while leaving the precision untouched. As can be
observed from the table, employing the rather expensive dependency features show
barely any improvement. Unexpectedly, the model that only uses word embedding
features is one of the best performing models. This was slightly
unexpected, since this is the model in which features are context
insensitive. However, the model that combines the word embeddings with
the word $n$-gram features outperforms the other linguistically
informed models that make use of lemmata, part-of-speech tags or
dependency information. The best performance is achieved by the model
that combines the word features, part-of-speech tags and the word
embeddings with an $F1$-score of 0.93 for animate words and 0.99 on
inanimate words. Adding more features does not result in any more
performance gain.

@@

The approach taken in this paper to create a model for animacy
classification using linguistically uninformed features proves to be
successful. We compared the performance of linguistically informed
models (using features such as part-of-speech and dependency tags) to
models that make use of lower dimensional representations of the
data. With the exception of the model that solely makes use of these
representations, all models benefit from adding these features. The
model that requires the least linguistic information (word $n$-grams
plus word embeddings) outperforms all linguistically informed
models. The best results are reported by the model that combines word
$n$-grams with Part-of-Speech $n$-grams and word embeddings.
